---
output: github_document
bibliography: vignettes/bibTeX/refs.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# SiGN

<!-- badges: start -->
<!-- badges: end -->

## Overview

SiGN is an R package that implements the Signals for Good News (SiGN) model to compute behavioural predictions in common operant choice paradigms.

## Installation

You can install the development version of SiGN from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("jpisklak/SiGN")
```

## Usage

```{r message = FALSE}
library(SiGN)

# Load built-in profile
kendall <- choice_params("kendall")

# Generate predictions
kendall_pred <- SiGN(kendall)

# View Results
kendall_pred$cp

# View intermediate calculations
kendall_pred$details
```

## Available Data and Profiles

The package includes:

- Built-in parameter profiles — Predefined setups for well-known procedures from the literature, including:

  - `"zentall"` [@Stagner_Zentall_2010]

  - `"kendall"` [@Kendall_1985]

  - `"fantino"` [@Fantino_1969]

- Fully vectorised parameter input — All model parameters can be customised and passed as vectors, enabling efficient simulation of multiple conditions.

- Built-in datasets — Two curated datasets from Dunn et al. (2024):

  - `subopt_full`: the complete dataset of suboptimal choice studies.

  - `subopt_avian`: a filtered subset focused on pigeons and starlings.

## References
